Hilbert space methods to approximate Gaussian processes using Stan

### 2019-03-21 Paper/manuscript/manuscript_05.tex

In this paper we focus on the basis function approximation via Laplace eigenfunctions for stationary covariance functions proposed by \citet{solin2018hilbert}. Basis function approaches behave computationally like linear models, which is an attractive property in modular probabilistic programming models where there is a big benefit id approximation specific computation is simple (Paul: What do you mean with the the last part of the sentence?). The Laplace eigenfunctions can be computed analytically and they are independent of the particular choice of the covariance kernel including the hyperparameters. While the pre-computation cost of the basis functions is $O(m^2n)$, the computational cost when only changing the covariance function parameters is $O(mn+m)$ (or even $O(m^3)$ if marginalization can be used). This is a big advantage in terms of speed for iterative algorithms such as Markov chain Monte Carlo (MCMC). Another advantage is the reduced memory requirements of automatic differentation methods used in modern probabilistic programming frameworks (cite!). This is because the memory requirements of automatic differentation rather scale with the computational complexity instead of with the usual memory requirements for the posterior density computation. The discussed basis function approach also provides an easy way to apply the non-centered parameterization of GPs, which reduces the posterior dependency between parameters representing the estimated function and the hyperparameters of the covariance function, which helps to further improve MCMC efficiency.

>>>>

In this paper we focus on the basis function approximation via Laplace eigenfunctions for stationary covariance functions proposed by \citet{solin2018hilbert}. Basis function approaches behave computationally like linear models, which is an attractive property in modular probabilistic programming models where there is a big benefit if approximation specific computation is simple (Paul: What do you mean with the the last part of the sentence?). The Laplace eigenfunctions can be computed analytically and they are independent of the particular choice of the covariance kernel including the hyperparameters. While the pre-computation cost of the basis functions is $O(m^2n)$, the computational cost when only changing the covariance function parameters is $O(mn+m)$.% (or even $O(m^3)$ if marginalization can be used).
This is a big advantage in terms of speed for iterative algorithms such as Markov chain Monte Carlo (MCMC). Another advantage is the reduced memory requirements of automatic differentation methods used in modern probabilistic programming frameworks (cite!). This is because the memory requirements of automatic differentation rather scale with the computational complexity instead of with the usual memory requirements for the posterior density computation. The basis function approach also provides an easy way to apply the non-centered parameterization of GPs, which reduces the posterior dependency between parameters representing the estimated function and the hyperparameters of the covariance function, which further improves MCMC efficiency.

\textcolor{red}{Aki: I moved this from section 2, needs to be combined with other intro}
We propose an approximate framework for fast and accurate inference
for Gaussian processes. Using a basis function expansion, we
approximate the Gaussian process with a linear model. This
representation has three main advantages: 1) it makes inference
considerably faster due to the linear structure, 2) it is simple to
implement, which makes it easy to use Gaussian processes as building
blocks in more complicated models and can be used as latent function
in non-Gaussian observational models allowing modelling flexibility,
3) it can be made arbitrary accurate and the trade-off between
computational complexity and approximation accuracy can easily be
controlled.
